\section{Core Module and Gateway Architecture}

The core module serves as the orchestration layer and entry point for security testing workflows. Research in this area focuses on architectural patterns, API gateway design, and microservice coordination for security tools.

\subsection{Microservice Architectures for Security Tools}

Modern security testing frameworks increasingly adopt microservice architectures to achieve modularity, scalability, and independent deployment of components. This approach allows different security testing modules to be developed, tested, and deployed independently while maintaining a cohesive system through well-defined APIs.

The core gateway acts as the single entry point for all requests, providing unified authentication, rate limiting, request routing, and response aggregation. This pattern ensures that security policies are consistently applied across all modules and simplifies the integration of new testing components.

\subsection{Input Validation and Normalization}

Robust input validation is critical for security testing tools to prevent injection attacks, ensure data integrity, and provide meaningful error messages. Research shows that proper URL normalization, protocol validation, and parameter sanitization at the gateway level significantly reduce downstream processing errors and security vulnerabilities.

Contemporary gateway implementations employ multi-layer validation strategies, including schema validation, type checking, range validation, and business logic validation. These layers work together to ensure that only well-formed, safe inputs propagate through the system.

\subsection{Request Orchestration and Workflow Management}

The core gateway coordinates complex multi-step security testing workflows by managing state, tracking progress, and ensuring proper sequencing of operations. Workflow orchestration patterns such as saga, choreography, and event-driven architectures enable resilient, fault-tolerant testing pipelines.

Research demonstrates that implementing proper circuit breakers, timeout management, and retry mechanisms at the gateway level significantly improves system reliability and prevents cascading failures when downstream services experience issues.
