\chapter*{Acknowledgments}
\addcontentsline{toc}{chapter}{Acknowledgments}

We would like to express our sincere gratitude to all those who have contributed to the successful completion of this mid-term report and the ongoing development of the Red Sentinel project.

First and foremost, we extend our deepest appreciation to our project supervisor for their invaluable guidance, continuous support, and constructive feedback throughout this research. Their expertise in web security and machine learning has been instrumental in shaping the direction of this work.

We are grateful to the Department of Electronics and Computer Engineering at Paschimanchal Campus, Tribhuvan University, for providing the necessary resources, infrastructure, and academic environment that enabled us to pursue this research.

We would like to acknowledge the open-source community for providing essential tools, frameworks, and datasets that formed the foundation of our system. Special thanks to the developers of TensorFlow, PyTorch, FastAPI, and the OWASP community for their contributions to web security research.

We also thank our families and friends for their unwavering support, patience, and encouragement throughout this journey. Their understanding during the challenging phases of this project has been invaluable.

Finally, we acknowledge the researchers whose prior work in automated vulnerability detection, transformer-based code generation, and adversarial machine learning inspired and informed our approach. Their contributions to the field have paved the way for innovations like Red Sentinel.

\vspace{1cm}

\begin{flushright}
Bipin Bhattarai \\
Seamoon Pandey \\
Sworup Bhandari \\
\vspace{0.5cm}
\textit{December, 2025}
\end{flushright}
